---
title: "Homework1_Instructions"
author: "Aidan O'Hara"
date: "2023-09-06"
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Note:

**1.** Please make sure that you use the given name for every question to ensure you get the full credit from the autograder, see this document as well as `hw1_basics.R` for correct names. Answer's with the wrong variable or function names will be ignored by GradeScope.

**2.** Please make sure your submission file name is `hw1_basics.R`.

**4.** Be sure to write in your name in the `YOUR NAME HERE` comment.

**4.** The due day for this assignment is **September 13, 2023 11:59 pm**, late submissions will not be accepted. 

# Vector
*do not use loops*

**1.** Create the following vector:\
**(a)** (1,2,3,...,19,20), Name `v1`;\
**(b)** (20,19,18,...,2,1), Name `v2`;\
**(c)** (1,3,5,...,17,19), Name `v3`;\
**(d)** (3,7,11,3,7,11,...,3,7,11) where there are 10 occurrences of 3, Name `v4`;\
**(e)** (3,7,11,3,7,11,...,3) where there are 11 occurrences of 3, 10 occurrences of 7 and 10 occurrences of 11, Name `v5`.\

**2.** Create a vector of the values of $e^xsin(x)$ at x = 3.0,3.1,3.2,...,6.0 Name `x1`;\

**3.** Calculate the following, name your answer `sum1`.
$$\sum^{100}_{i=10}(i^3+4i^2)$$

**4.** Use the function `paste` to create the following character vectors of length 30:\
**(a).** ("label 1", "label 2", ....., "label 30").
Notice: there is a single space between 'label' and number following, Name `str1`;\
**(b).** ("function1", "function2", ..., "function30").
In this case, there is no space between 'function' and number following, Name `str2`;\

# Matrix

**1.** Suppose: 
$$
A = \begin{bmatrix}
1 & 1 & 3 \\
5 & 2 & 6 \\
-2 & -1 & -3
\end{bmatrix}
$$

**(a)** Check the $A^3 = 0$ where 0 is a $3 \times 3$ matrix with every entry equal to 0.
Name `neo`\
**(b)** Replace the third column of $A$ by the sum of the second and third columns.
Name `neo2`;

**2.** 
$$
B = \begin{bmatrix}
10 & -10 & 10 \\
10 & -10 & 10 \\
... & ... & ...\\
10 & -10 & 10
\end{bmatrix}
$$ 
\
Calculate and save the $3 \times 3$ matrix $B^t B$ (look at the help for the `crossprod` R function for additional guidance) Name `redOrBlue`;\

**3.** Solve the following system of linear equations in five unknowns by considering an appropriate matrix equation $Ax = y$.\
$$
\begin{aligned}
x_1 + 2x_2 + 3x_3 + 4x_4 + 5x_5 &= 7\\
2x_1 + x_2 + 2x_3 + 3x_4 + 4x_5 &= -1\\
3x_1 + 2x_2 + x_3 + 2x_4 + 3x_5 &= -3\\
4x_1 + 3x_2 + 2x_3 + x_4 + 2x_5 &= 5\\
5x_1 + 4x_2 + 3x_3 + 2x_4 + x_5 &= 17
\end{aligned}
$$\

Make use of the special form of the matrix A.
The method used for the solution should easily generalize to a larger set of equations where the matrix A has the same structure; hence the solution should not involve typing in every number of A.
*Use `solve` function.* Name: `smith`;

**4.** Create a 6 × 10 matrix of random integers chosen from 1, 2,..., 10 by executing the following two lines of code:

```{r,eval=FALSE, echo=TRUE}
set.seed(75)
aMat <- matrix( sample(10, size=60, replace=T), nr=6)
```

**(a)** Find the number of entries in each row which are greater than 4.
Name `fours`;\
**(b)** Which rows contain exactly two occurrences of the number seven?
Name `seven`;\
**(c)** Find those pairs of columns whose total (over both columns) is greater than 75.
The answer should be a matrix with two columns; so, form example, the row (1, 2) in the output matrix means that the sum of columns 1 and 2 in the original matrix is greater than 75.
Repeating a column is permitted; so, for example, the final output matrix could contain the rows (1, 2), (2, 1) and (2, 2).
Name `counts`;\

**5.** Calculate the following and assign them to the given variable name *do not use loops*:\
**(a)** `somewhere`\
$$
\sum_{i = 1}^{20}\sum_{j = 1}^{5}\frac{i^4}{(3+j)}
$$\
**(b)** `something`\
$$
\sum_{i = 1}^{20}\sum_{j = 1}^{5}\frac{i^4}{(3+ij)}
$$\
**(c)** `somebody`\
$$
\sum_{i = 1}^{10}\sum_{j = 1}^{i}\frac{i^4}{(3+ij)}
$$\

# Data Frame

**1.** Write a function, called `myListFn`, which takes a single argument n and implements the following algorithm:

1.  Simulate `n` independent numbers, denoted $\textbf{x} = (x_1,x_2,...,x_n)$, from the N(0,1) distribution.
2.  Calculate the mean, $\bar{x}$.
3.  If $\bar{x} > 0$, simulate `n` independent numbers, denoted $\textbf{y} = (y_1,y_2,...,y_n)$, from the exponential density with mean $\bar{x}$. If $\bar{x} < 0$ then simulate *n* independent numbers, denoted $\textbf{z} = (z_1,z_2,...,z_n)$ from the exponential density with mean $-\bar{x}$. Set $\textbf{y} = (y_1,y_2,...,y_n) = -\textbf{y}$.
4.  Calculate *k* which is the number of *j* with $|y_j| > |{x_j}|$.
5.  Return the list of **x**, **y**, and *k* with names `xVec`, `yVec`, and `count` respectively.

**2.** In order to test the functions in this question, you will need an `array`.
We can create a three dimensional test `array` as follows:\
`testArray <- array( sample( 1:60, 60, replace=F), dim=c(5,4,3) )`\

The above line creates a 5 × 4 × 3 array of integers which can be represented in mathematics by:\
$\{x_{i,j,k}: i = 1,2,...,5; j = 1,2,3,4; k = 1,2,3\}$\

Note that `apply(testArray, 3, tmpFn)` mean that the index *k* is retained in the answer and the function `tmpFn` is applied to the 3 matrices:\
$\{x_{i,j,1}: 1 \le i \le 5; 1 \le j \le 4\}$,\
$\{x_{i,j,2}: 1 \le i \le 5; 1 \le j \le 4\}$ and $\{x_{i,j,1}: 1 \le i \le 5; 1 \le j \le 4\}$.\
Similarly `apply(testArray, c(1,3), tmpFn)` means that indices *i* and *k* are retained in the answer and the function `tmpFn` is applied to 15 vectors: $\{x_{1,j,1}:1 \le j \le 4\}$, $\{x_{2,j,1}:1 \le j \le 4\}$, etc.\
The expression `apply(testArray, c(3,1), tmpFn)` does the same calculation but the format of the answer is different: when using `apply` in this manner, it is always worth writing a small example in order to check that the format of the output of `apply` is as you expect.\

**(a)** Write a function `testFn` which takes a single argument which is a 3-dimensional array.
If this array is denoted $\{x_{i,j,k}: i = 1,2...,d_1; j=1,2,...,d_2; k = 1,2,...,d_3\}$, then the function `testFn` returns a list of the $d_1 \times d_2 \times d_3$ matrix $\{w_{i,j,k}\}$ and the $d_2 \times d_3$ matrix $\{z_{j,k}\}$ where:

$w_{i,j,k} = x_{i,j,k} - \min_{i = 1}^{d_1}x_{i,j,k}$ and $z_{j,k} = \sum_{i=1}^{d_1}x_{i,j,k} - \max_{i = 1}^{d_1}x_{i,j,k}$

**(b)** Now suppose we want a function `testFn2` which returns the $d_2 \times d_3$ matrix $\{z_{j,k}\}$ where\
$z_{j,k} = \sum_{i=1}^{d_1}x_{i,j,k}^k$

**3.** Load the `penguins` data set from `library(palmerpenguins)`.
Copy the data to a new variable `peng` and then create the following new columns: - `bill_depth_in`: bill depth in inches.
- `bill_length_in`: bill length in inches.
- `flipper_length_in`: flipper length in inches.
- `body_mass_kg`: body mass in kilograms - `body_mass_lb`: body mass in pounds

Convert mm to in by using a factor of `0.0393701`, and kg to lb using a factor of `2.20462`.\

**4.** Create three subsets of `peng`, called `pengPres1`, `pengPres2`, `pengPres3`.
Each subset should include only Gentoo, Chinstrap, and Adelie species respectively.
Only include `flipper_length_in` and `body_mass_lb`, omit any rows with `NA` values.\

**5.** Remove any rows from `peng` that contain `NA` values.
Construct `groupPengs` to contain 100 rows about samples of 5 rows from `peng`.
Populate your rows with the following columns:

-   `groupMass_g`: combined mass in grams of all the penguins in the sample
-   `longest_beak_mm`: the longest beak in mm recorded in the sample
-   `happy`: a binary indicating if all 5 penguins are the same species
-   `wellFed`: a binary indicating if the the mean of the sample is greater than the mean mass of penguins from `peng`

# Function

**1.** Create two functions `function1` and `function2`,

**(a).** `function1(xv)` return the vector $(x_1,x_2^2,...,x_n^n)$, use the vector $xv = (0.0,0.1,...,0.8,0.9,1.0)$ to calculate `func1_ans`;\

**(b).** `function2(xv)` return $(x_1,\frac{x_2^2}{2},\frac{x_3^3}{3},...,\frac{x_n^n}{n})$, use the vector $xv = (0.0,0.1,...,0.8,0.9,1.0)$ to calculate `func2_ans`;\

**(c).** Write a function `function3` which takes 2 arguments x and n where x is a single number and n is a strictly positive integer.
The function should return the value of\
$$
1 + \frac{x}{1} + \frac{x^2}{2} + \frac{x^3}{3} + ... + \frac{x^n}{n}
$$ 
\
then use the vector $xv = (0.0,0.1,...,0.8,0.9,1.0)$ to calculate `func3_ans`;\

**2.** Write a function called `findodd(x)` that returns a list of the odd numbers between 1 and input `x`.

**3.** Write a function that receives two inputs, a number `x`, and another number `y`, the function should return `x` or the next largest number divisible by `y`.
*hint: use the modulo operator `%%.`*

Call your function `modNumber(x,y)`.

*ex: `modNumber(50,16)` should return `[1] 64` and `modNumber(64,16)` should also return `[1] 64`*

*Your supplied `modNumber(x,y)` should return `[1] 64` and `modNumber(64,16)` should also return `[1] 64`.*
\newpage

**4.** Write a function, `matFn`, which takes 2 arguments *n* and *k* which are positive integers.
It should return the *n × n* matrix: \
$$
\begin{bmatrix}
k & 1 & 0 & 0 & ... & 0 & 0 \\
1 & k & 1 & 0 & ... & 0 & 0 \\
0 & 1 & k & 1 & ... & 0 & 0 \\
0 & 0 & 1 & k & ... & 0 & 0 \\
... & ... & ... & ... & ... & ... & ... \\
0 & 0 & 0 & 0 & ... & k & 1 \\
0 & 0 & 0 & 0 & ... & 1 & k
\end{bmatrix}
$$ 

*Hint: First try to do it for a specific case such as n = 5 and k = 2 on the Command Line.*\


**5.** Zeller's congruence is the formula: $f = ([2.6m - 0.2] + k + y + [y/4] + [c/4] - 2c) mod 7$ where $[x]$ denotes the integer part of $x$; for example $[7.5] = 7$.
Zeller's congruence returns the day of the week $f$ given: $k$ = the day of the month,\
$y$ = the year in the century\
$c$ = the first 2 digits of the year (the century number)\
$m$ = the month o=number (where January is month 11 of the preceding year, February is month 12 of the preceding year, March is month 1, etc.)\
For example, the date 21/07/1963 has m = 5, k = 21, c = 19, y = 63; whilst the date 21/2/1963 has m = 12, k = 21, c = 19 and y = 62.

Write a function `weekday(day,month,year)` which returns the day of the week when given the numerical inputs of the day, month and year.
Note that the value of 1 for $f$ denotes Sunday, 2 denotes Monday, etc.\

# More Function

**1.**
**(a)** Write a function `quadmap(start, rho, niter)` which returns the vector $(x_1,...,x_n)$ where $rx_{k-1}(1-x_{k-1})$ and\
  - `niter` denotes $n$,  
  - `start` denotes $x_1$  
  - `rho` denotes $r$.\

Try out the function you have written: - for \$ r=2 \$ and $0<x_1<1$ you should get $x_n \to 0.5$ as $n \to \infty$.
- try `tmp <- quadmap(start = 0.95, rho=2.99, niter=500)`, try `plot(tmp, type="l")` in console.\

**(b)** Now write a function, `quad2`, which determines the number of iterations needed to get \$\|x_n - x\_{n-1}\|\<0.02.
So this function has only 2 arguments: `start` and `rho` (For `start=0.95` and `rho=2.99`, the answer is 84.)\

**2.**
**(a)** Suppose `matA` is a matrix containing some occurrences of `NA`.
Pick out the submatrix which consists of all columns which contain no occurrence of `NA`.
So the objective is to write a function which takes a single argument which can be assumed to be a matrix and returns a matrix.\

**(b)** Now write a function which takes a single argument which can be assumed to be a matrix and returns the submatrix which is obtained by deleting every row and column from the input matrix which contains an `NA`.\

**3.**
Experiment with different ways of defining a function which calculates the following double sum for any value of $n$.\
$$
f(n)=\sum_{i = l}^{n}\sum_{s=l}^{r}\frac{s^2}{10 + 4r^3}
$$\
**(a)** First use a loop within a loop.\

**(b)** Write a function `funB` which uses the functions row and col to construct a matrix with suitable entries so that the sum of the matrix gives the required answer.\

**(c)** Write a function `funC` which uses the function outer to construct a matrix with suitable entries so that the sum of the matrix gives the required answer.\

**(d)** Create a function which takes a single argument `r` and calculates\
$$
\sum_{s=l}^{r}\frac{s^2}{10 + 4r^3}
$$ 
\
Then write a function `funD` which uses `sapply` to calculate the double sum.
Note that `sapply` is just a combination of `unlist` and `lapply`.
Is there any increase in speed gained by using this information (`funE`)?

**(e)** Write a function which takes two arguments `r` and `s` and calculates\
$$
\frac{I(s \le r)s^2}{10 + 4r^3}
$$ where $I$ denotes the indicator function.
Then write a function `funF` which calculates the double sum by using mapply to calculate all the individual terms.\

**(f)** Determine which function is fastest.
Save your fastest function as `fastestFun`.\

**4.** The waiting time of the $n^th$ customer in a single server queue.
Suppose customers labelled $C_0, C_1,...,C_n$ arrive at times $\tau = 0, \tau_1, . . . , \tau_n$ for service by a single server.
The inter-arrival times $A_1 = \tau_1 - _0,...,A_n = \tau_n - \tau_{n-1}$ are independent and identically distributed random variables with the exponential density\
$$
\lambda_ae^{-\lambda_ax} \text{ for } x \ge 0
$$\
The service times $S_0, S_1,..., S_n$ are independent and identically distributed random variables which are also independent of the inter-arrival times with the exponential density\
$$
\lambda_se^{-\lambda_sx} \text{ for } x \ge 0
$$\
Let $W_j$ denote the waiting time of customer $C_j$ .
Hence customer $C_j$ leaves at time $\tau_j + W_j + S_j$.
If this time is greater than $\tau_{j+1}$ then the next customer, $C_{j+1}$ must wait for the time $\tau_j + W_j + S_j - \tau_{j+1}$.
Hence we have the recurrent relation $W_0 = 0$, $W_{j+1} = \max\{0,W_j + S_j - A_{j+1}\} \text{ for } j = 0,1,...,n-1$\

Write a function `queue(n, aRate, sRate)` which simulates one outcome of $W_n$ where `aRate` denotes $\lambda_a$ and `sRate` denotes $\lambda_s$.
Try out your function on an example such as `queue(50,2,2)`.\

**5.** A random walk.
A symmetric simple random walk starting at the origin is defined as follows.
Suppose $X_1, X_2,...$ are independent and identically distributed random variables with the distribution: $+1 \text{ with probability } 1/2$ or $-1 \text{ with probability } 1/2$\

Define the sequence $\{S_n\}_{n \ge 0}$ by\

$$
S_0 = 0 \\
S_n = S_{n-1} + X_n \text{ for } n = 1,2,...
$$\
Then $\{S_n\}_{n \ge 0}$ is a symmetric simple random walk starting at the origin.
Note that the position of the walk at time $n$ is just the sum of the previous $n$ steps: $S_n = X_1 +...+ X_n$.\

**(a)** Write a function `rwalk(n)` which takes a single argument n and returns a vector which is a realization of $(S_0, S_1,..., S_n)$, the first $n$ positions of a symmetric random walk starting at the origin.
*Hint: the code `sample(c(-1,1), n, replace=TRUE, prob=c(0.5,0.5))` simulates* $n$ steps\

**(b)** Now write a function `rwalkPos(n)` which simulates one occurrence of the walk which lasts for a length of time $n$ and then returns the length of time the walk spends above the x-axis.
*(Note that a walk with length 6 and vertices at 0, 1, 0, -1, 0, 1, 0 spends 4 units of time above the axis and 2 units of time below the axis.)*\

**(c)** Now suppose we wish to investigate the distribution of the time the walk spends above the x-axis.
This means we need a large number of replications of `rwalkPos(n)`.
Write two functions: `rwalkPos1(nReps,n)` which uses a loop and `rwalkPos2(nReps,n)` which uses replicate or `sapply`.
Compare the execution times of these two functions.\

**(d)** In the previous question on the waiting time in a queue, a very substantial increase was obtained by using a vector approach.
Is that possible in this case?
